\section{Discussion and limitations}
\label{sec:discussion}

\paragraph{What the schedule buys, and what it does not.} Across MLPs and a
small transformer, the picture is consistent: cheap front steps carry most
of the per-step progress; deep credit assignment is needed occasionally, in
\emph{consecutive} doses, at moments a loss plateau identifies. At
${\sim}4$--$6\%$ full-depth budgets, \fw{} recovers roughly two-thirds of the
front-to-full accuracy gap on MLPs and outperforms or matches every fixed
schedule we tested (Sections~\ref{sec:exp2}, \ref{sec:exp3})---but it does
not match always-full training
($1.3$--$3.3$ points on MLPs; a larger gap in validation loss on the
transformer, where it recovers about a fifth of the front-to-full gap). \fw{}
is a compute-quality trade, not a free lunch, and the trade steepens with
task difficulty in both directions: adaptivity matters more, and full depth
remains further out of reach. The study's organizing observation is about
variance across conditions, not means within one: every fixed alternative
we tested owns at most one winning cell and at least one failing one,
because each hard-codes an assumption---an early budget (\textsc{ftlp}), a
freeze time (\textsc{ratchet}), a cadence (\textsc{pburst},
\textsc{periodic}), a single switch (\lpft{})---that some tested condition
violates, stationary conditions included. What \textsc{full} actually
sells is unconditional robustness; \fw{} is the cheap schedule that comes
closest to inheriting it.

\paragraph{What we believe transfers.} Four mechanisms, not the numbers.
(i)~\emph{Nested beats partitioned}: depth truncation is exclusive to the
top block, so compute-aware schedules should treat burst timing and block
choice as separate decisions---and our ablation shows the timing is the
active ingredient. (ii)~\emph{Burst-as-measurement with backoff}: any
scheduler that must decide whether expensive credit assignment is still
productive faces the cried-wolf problem; exponential backoff is a simple,
robust answer. (iii)~\emph{Isolated deep steps are dangerous}: interleaved
partial-training schemes need burst structure, and under Adam neither
gradient clipping nor regularization substitutes for it. (iv)~\emph{The
optimizer fires with the BP, gently}: moment estimates are part of the
credit-assignment machinery, should not outlive inter-burst gaps, and must
re-enter with a warmup---fresh state at full aggression is worse than stale
state; the episodic formulation also removes trunk optimizer memory for the
vast majority of steps.

\paragraph{What we do not claim.} All experiments are small (MLPs; a 10.8M
transformer on 1MB of text) with three seeds; absolute numbers are far from
competitive by design. We have not tested \fw{} at scale, under distribution
shift, with pretrained initializations, or against learned-schedule
baselines; the front was always the topmost block, and other cheap subspaces
(low-rank, sparse) are untested. The origins study
(Section~\ref{sec:exp1}) is scoped by its own depth result: local wiring
rules match dense training only in shallow networks.

\paragraph{Open directions.} Making the trigger's implicit estimate of
inequality~\eqref{eq:greedy} explicit (cheap stochastic probes of
$\|\nabla L\|$) would turn \fw{} into an approximate optimal stopping rule.
The episodic-optimizer ablation suggests a broader design space:
per-burst warmup lengths, moment half-lives matched to burst spacing, and
trust-ratio update clipping as the Adam-compatible safety net. At scale, the
memory argument sharpens---trunk optimizer state that exists only ${\sim}6\%$ of
the time is a different systems regime---and the natural next testbed is a
pretrained model, where \fw{} becomes an adaptive fine-tuning schedule and
\lpft{} is the one-switch special case it generalizes.

\section{Conclusion}
Most training steps do not need deep credit assignment; the hard part is
knowing which ones do---and every cheap schedule is an answer to that
question. The fixed answers each win where their built-in assumption holds
and fail outright where it does not: the best mid-budget schedule in this
study (deep-first, right budget) is also its worst cell (deep-first, wrong
budget) and drops to the never-deep floor when the data changes. \fw{}'s
answer---front steps by default, consecutive bursts when a plateau says the
cheap subspace is exhausted, backoff when a burst reveals the plateau was
global---carries no assumption to violate: one configuration, two
architectures, two optimizers, stationary and shifted data, never worse
than second-best among cheap schedules and never below the never-deep
floor, at $20$--$25\%$ of full backprop's compute. Where schedules fail,
the failures localize to mechanisms we could measure (isolated-step churn,
stale optimizer moments, irrevocable early commitments) and repair (bursts,
episodic state, revocability). Credit assignment is a budgeted resource;
\fw{} is a first controller for spending it---its backoff cap is the dial
that chooses the budget, and its robustness, not its peak, is the point.

\paragraph{Reproducibility.} Code, run logs, and all raw result CSVs are
available at \url{https://github.com/XTAM-AI/front-wheeler}. Every number in
this paper is the mean over seeds $\{0,1,2\}$ with the exact commands
recorded in the run scripts.
