\section{Front wheeler on a transformer}
\label{sec:exp3}

% TODO-final-numbers: seeds 1,2 pending; all numbers below are seed 0 and
% will be replaced by 3-seed means when the suite completes.

\paragraph{Protocol.} Character-level language modeling on tiny
Shakespeare: a 10.8M-parameter decoder-only transformer (6 layers,
$d{=}384$, 6 heads, context 256, dropout $0.2$), AdamW ($3{\times}10^{-4}$,
200-step warmup, cosine decay, weight decay $0.1$, clip $1.0$), 5{,}000
steps, batch 64, 3 seeds. The front is the last transformer block, the final
LayerNorm, and the LM head; the LM head is \emph{untied} from the token
embedding so front steps cannot leak trunk updates through weight tying.
Front steps run the trunk under \texttt{no\_grad}. Schedules as in
Section~\ref{sec:exp2}; metric is validation loss.

\paragraph{Only adaptive bursts convert a small budget into progress.}
At seed 0: full BP reaches $1.50$; front-only $1.84$. Between those anchors,
the fixed schedules barely move: \lpft{} at a $10\%$ budget gains $0.004$
over front-only, and periodic interleaving is \emph{destructive}---$2.04$ at
a $10\%$ budget and $2.44$ at $20\%$, far worse than doing nothing deep.
\fw{} at a ${\sim}6\%$ budget reaches $1.76$, matching \lpft{} at a $50\%$
budget---an ${\sim}8\times$ budget efficiency over the one-switch
schedule---and \bcd{} tracks it at $1.79$. The gap to always-full remains
substantial (\fw{} recovers ${\sim}22\%$ of the front-to-full gap, vs.\
${\sim}$two-thirds on MLPs): deep trunks need more deep training on harder
tasks, and \fw{} trades quality for compute rather than matching full BP.

\paragraph{Interleaving toxicity is optimizer-dependent and clipping does
not save Adam.} The periodic failure is the MLP collapse of
Section~\ref{sec:exp2} in a new guise, with two differences. First, damage
scales monotonically with interleaving \emph{frequency} (worse at period 5
than 50), implicating repeated perturbation---each isolated deep step moves
trunk features under a head sharply co-adapted to the old ones, and the
system churns instead of converging---rather than any single bad step.
Second, it happens \emph{despite} gradient clipping, which cured the SGD
case: Adam's update $m/\sqrt{v}$ is approximately scale-invariant in the
gradient, so clipping the gradient does not clip the step. Trunk moment
estimates also go stale across gaps ($\beta_2 = 0.999$ means a 50-step burst
refreshes only ${\sim}5\%$ of $v$), the schedule-induced analogue of the
sparse-update pathology that motivates SparseAdam for embedding tables.

\paragraph{Episodic optimizer state: the optimizer fires with the BP.}
If moment estimates are part of the credit-assignment machinery, their
\emph{state} should be episodic like the bursts themselves. We test four
variants: \textsc{persist} (status quo), \textsc{reset} (trunk moments
cleared at burst entry), \textsc{reset-warmup} (reset plus a 10-step
in-burst learning-rate warmup, mirroring fine-tuning practice when
unfreezing), and \textsc{episodic} (Adam state exists only inside bursts;
front phases run on SGD). Beyond stability, \textsc{episodic} has a systems
consequence: optimizer memory for the trunk (two fp32 moments per parameter)
need exist only during the ${\sim}4$--$6\%$ of steps spent in bursts.
% TODO-final-numbers: ablation table ({persist,reset,reset-warmup,episodic}
% x {fw, periodic-10} x 3 seeds) pending; discriminates moment-staleness
% from co-adaptation churn (reset rescues periodic iff staleness dominates).
