\section{Introduction}

A full backward pass costs roughly twice the forward pass, must store
activations to full depth, and touches every parameter on every step. Much of
that work may be wasted: late layers can be updated by backpropagating
through only the final two weight matrices, and for long stretches of
training those cheap updates are all that is improving the loss. The question
is \emph{when} the network actually needs deep credit assignment---and
whether that decision can be made online, from signals training already
produces.

We propose \fw{} (\emph{front wheeler}). By default every step is a
\emph{front} step: the trunk runs forward without building a graph, and only
the last hidden layer and readout are updated. A plateau detector watches an
EMA of the training loss; when improvement stalls, \fw{} fires a \emph{burst}
of consecutive full-backpropagation steps, then returns to front mode. An
exponential backoff handles the ``boy who cried wolf'' failure: if a burst
itself yields no improvement, the plateau is global rather than the front's
fault, and the evidence bar for the next alarm doubles. The schedule is a
closed-loop controller---a cost-aware Gauss--Southwell rule over
\emph{nested} parameter blocks, where each burst doubles as a measurement of
whether depth is still worth paying for (Section~\ref{sec:framework}).

\paragraph{Contributions.} On MLPs (MNIST, CIFAR-10) and a 10.8M-parameter
character-level transformer, three seeds throughout:
\begin{enumerate}
  \item \textbf{Adaptive bursts beat or match every fixed alternative at
  low budget.}
  At a matched ${\sim}4$--$5\%$ full-backprop budget, \fw{} beats periodic
  schedules by $+1.5$ (MNIST) and $+5.1$ (CIFAR-10) points and one-switch
  \lpft{} \citep{kumar2022finetuning} by $+2.0$ and $+6.1$---margins that
  \emph{grow} with task difficulty, and that survive the worst rather than
  best trigger threshold. On the transformer, \fw{} spending ${\sim}6\%$ of
  steps at full depth matches \lpft{} spending $50\%$, and only the trigger
  family (\fw{} and its \bcd{} ablation) improves at low budget on never
  back-propagating deep at all. The backoff cap doubles as a \emph{budget
  dial}: one parameter traces a monotone Pareto curve ($6$--$49\%$
  budgets) dominating \lpft{} and periodic at every tested budget on the
  transformer. A progressive-freezing ratchet \citep{brock2017freezeout,
  liu2021autofreeze} ties \fw{} on MNIST at matched backward compute, but
  its freeze timing is task-critical and irrevocable: the same setting is
  catastrophic on the transformer, while \fw{} runs identical trigger
  hyperparameters throughout. Controls disentangle the mechanism
  (Section~\ref{sec:controls}): a deep-first schedule owns both the best
  cheap stationary cell (at the right budget) and the worst cell in the
  study (at \fw{}'s budget), and falls to the never-deep floor under a
  mid-training distribution shift, while an oracle-period clock ties or
  slightly trails \fw{} everywhere---bursts carry the performance; the
  trigger's product is finding the schedule without the oracle, and
  robustness when assumptions break.
  \item \textbf{Isolated deep steps are dangerous; bursts make them safe.}
  Full-depth steps sprinkled one at a time into cheap training collapse MLPs
  (dead-ReLU cascade driven by error signals from a co-adapted head, measured
  growing to ${\sim}20\times$ their initial scale) and degrade the
  transformer \emph{monotonically with interleaving frequency}: periodic
  interleaving at a $20\%$ budget trains to a worse loss than spending
  nothing on depth at all. Consecutive phases (\fw{} bursts, \lpft{}'s single
  switch) do not exhibit the failure.
  \item \textbf{The optimizer should fire with the backpropagation.} Under
  SGD the collapse is prevented by gradient clipping; under Adam it is not,
  because Adam's scale-invariant update neutralizes clipping, and trunk
  moment estimates go stale across inter-burst gaps (the schedule-induced
  analogue of the sparse-embedding pathology that motivates SparseAdam). We
  formulate and test \emph{episodic} optimizer state---moments that exist
  only within bursts. \fw{} itself converges identically with or without
  the reset (its bursts re-equilibrate the trunk organically), so episodic
  state is a free systems win---the trunk's optimizer memory vanishes for
  the ${\sim}94\%$ of steps spent in front mode---rather than a
  convergence requirement. \fw{} saves compute time and average memory
  traffic; peak activation memory is set by the deepest step and is
  unchanged, with burst-scoped activation checkpointing the natural path to
  reducing it.
  \item \textbf{The trigger, not the block, carries the benefit.} A
  block-coordinate ablation (\bcd{}: identical trigger, bursts update the
  complementary block) tracks \fw{} closely on all testbeds, isolating the
  closed-loop schedule as the active ingredient---while only the nested
  scheme converts cheap phases into compute savings, since trunk-only steps
  still backpropagate activation gradients through the head.
\end{enumerate}

Appendix~\ref{sec:exp1} reports the experiments that motivated \fw{}: on
frozen random networks, connectivity search alone nearly matches weight
training, and a trained readout on frozen features already recovers most of
the accuracy---evidence that the bulk of per-step improvement does not
require deep credit assignment. We emphasize scope throughout: all results
are deliberately small-scale; we characterize the schedule, not the state of
the art. Code and all run logs are released.
