\section{Related work}

\paragraph{Scheduling and partial credit assignment.} Linear probing
followed by fine-tuning (\lpft{}) \citep{kumar2022finetuning} is the
one-switch special case of \fw{}, motivated in the transfer setting by
feature distortion: a fresh head sends large, misaligned gradients into good
pretrained features. Our stability findings are the from-scratch, per-step
version of the same mechanism. Closest to \fw{} is the adaptive
layer-freezing line: FreezeOut \citep{brock2017freezeout} anneals early
layers out on a fixed schedule, while AutoFreeze \citep{liu2021autofreeze},
Egeria \citep{wang2023egeria}, and SmartFRZ \citep{li2023smartfrz} freeze
layers adaptively from convergence signals (gradient norms, plasticity,
learned predictors). These are \emph{ratchets}: layers freeze bottom-up,
progressively, and (near-)permanently, saving on the order of
$20$--$50\%$ of training FLOPs. \fw{} inverts the default and changes the
structure: the trunk is frozen from the start, deep credit assignment
happens in \emph{temporary, revocable bursts}, and the operating point is a
$4$--$6\%$ full-depth budget rather than a moderate discount. Our
isolated-step toxicity result is complementary to this literature: it
explains why the ratchet structure is safe (frozen layers are never
perturbed by sporadic deep steps) and why naive interleaving---the natural
first alternative to a ratchet---is not. Decoupled training replaces the
global backward pass with learned local surrogates
\citep{jaderberg2017decoupled} or greedy layer-wise objectives
\citep{belilovsky2019greedy}. Fixed schedules and ratchets alike cannot
express \fw{}'s backoff, which handles unproductive expensive phases by
raising the evidence bar for the next one. Block
coordinate descent with greedy (Gauss--Southwell) selection
\citep{nutini2015coordinate} is the classical frame for adaptive block
updates; its theory assumes uniform per-block costs, whereas the cheap block
here is the \emph{only} one admitting depth truncation
(Section~\ref{sec:framework})---the cost asymmetry is the problem statement.

\paragraph{Supermasks and random features (origins).} \citet{frankle2019lottery}
showed dense networks contain sparse trainable subnetworks;
\citet{zhou2019deconstructing} and \citet{ramanujan2020whats} showed a mask
over \emph{untrained} weights already performs well, with existence
guarantees by \citet{malach2020proving} and \citet{pensia2020optimal}, and
\citet{gaier2019weight} demonstrated weight-agnostic architectures. Our
origins study (Section~\ref{sec:exp1}) transplants local learning
rules---feedback alignment \citep{lillicrap2016random}, direct feedback
alignment \citep{nokland2016direct}, three-factor Hebbian rules
\citep{fremaux2016neuromodulated, williams1992simple}; see also
forward-forward \citep{hinton2022forward}---from weight updates to
mask-score updates via straight-through estimation
\citep{bengio2013estimating}. Its readout control, a frozen trunk with a
trained top, is what motivated \fw{}'s premise that cheap top-only updates
carry most per-step progress.
