\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\graphicspath{{figures/}}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{caption}
\usepackage[colorlinks=true,linkcolor=blue!60!black,citecolor=blue!60!black,urlcolor=blue!60!black]{hyperref}

\newcommand{\ep}{\textsc{ep}}
\newcommand{\dfa}{\textsc{dfa}}
\newcommand{\rmhebb}{\textsc{rm-hebb}}
\newcommand{\fw}{\textsc{fw}}
\newcommand{\lpft}{\textsc{lp-ft}}
\newcommand{\bcd}{\textsc{bcd}}
\newcommand{\thh}{\theta_{\mathrm{h}}}
\newcommand{\tht}{\theta_{\mathrm{t}}}

% TODO alternative titles:
%  "Front Wheeler: Plateau-Triggered Backpropagation"
%  "Paying for Depth Only When It Pays: Adaptive Bursts of Backpropagation"
\title{Front Wheeler: Training with Full Backpropagation\\
Only When It Pays}

\author{Mamdouh Alramadan\\
XTAM Research Labs\\
\texttt{research@xtam.ai}}

\date{}

\begin{document}
\maketitle

\begin{abstract}
Backpropagating through every layer on every step is the single largest cost
of training, yet most steps may not need deep credit assignment. We propose
\emph{front wheeler} (\fw{}): by default, each step updates only the last
hidden layer and the readout (backward through two weight matrices); a
\emph{burst} of full-backpropagation steps fires only when a loss plateau
signals that the cheap subspace is exhausted, with exponential backoff when a
burst itself proves unproductive. On MLPs (MNIST, CIFAR-10), \fw{} at a
${\sim}4\%$ full-backprop budget beats matched periodic schedules and
one-switch linear-probe-then-fine-tune (\lpft{}) by margins that grow with
task difficulty ($+1.5$ to $+6.1$ points). On a 10.8M-parameter transformer
language model, \fw{} with ${\sim}6\%$ of steps at full depth matches
\lpft{} spending $50\%$, and only the trigger-driven schedules improve at low budget on never
back-propagating deep at all; a single backoff parameter further
acts as a budget dial, tracing a monotone compute--quality curve
($6$--$49\%$ budgets) that dominates one-switch and periodic schedules at
every tested budget on the transformer. Under a mid-training distribution
shift, \fw{} is the best cheap schedule, invariant to when the shift
occurs, while the deep-first schedule that wins one stationary budget falls
to the never-deep floor. Across all conditions the pattern is consistent:
every fixed alternative is a bet on one assumption---a budget, a freeze
time, a clock---and fails outright when it is wrong, on stationary tasks as
much as shifted ones; \fw{}, with one configuration throughout, is never
worse than second-best among cheap schedules and never fails. Three mechanism findings accompany the
method: (i) \emph{isolated} full-depth steps interleaved with cheap steps
are actively destructive---collapsing MLPs via dead units and degrading the
transformer monotonically with interleaving frequency---and burst structure
is what makes deep steps safe; (ii) under Adam the damage persists despite
gradient clipping, since Adam's scale-invariant updates neutralize clipping,
motivating an \emph{episodic} optimizer whose moment estimates do not
outlive inter-burst gaps; (iii) a block-coordinate ablation shows the
closed-loop trigger, not the block structure, carries most of the benefit.
Experiments are small-scale by design; we characterize the schedule, not the
state of the art.
\end{abstract}

\input{sections/intro}
\input{sections/glossary}
\input{sections/related}
\input{sections/framework}
\input{sections/exp2}
\input{sections/exp3}
\input{sections/exp1}
\input{sections/discussion}

\section*{Acknowledgments}
This work was carried out at XTAM Research Labs. Transparency about tooling:
this work was AI-assisted (Anthropic's Claude---Claude Code, Claude Fable
5---for implementation, experiment orchestration, debugging, and drafting),
under the direction and review of the author. All results were produced by
the released code, and every reported number was checked against the raw run
logs in the repository.

\bibliographystyle{unsrtnat}
\bibliography{refs}

\appendix

\section{Experimental infrastructure}
\label{app:infra}
All experiments ran on a single workstation: one NVIDIA RTX PRO 6000
Blackwell (96\,GB), 64 CPU cores, 256\,GB RAM; PyTorch 2.13 / CUDA~13 in a
micromamba environment. The complete experimental suite---both MLP studies,
the transformer study, and all ablations and sweeps, three seeds
throughout---consumes under ten GPU-hours end to end. Datasets: MNIST
\citep{lecun1998gradient}, CIFAR-10 \citep{krizhevsky2009learning}, and the
tiny Shakespeare corpus. Exact commands for every run are recorded in the
repository's \texttt{scripts/} directory; per-run CSV logs (including
wall-clock times) are in \texttt{results/}.

\section{Two failure modes of mask-search rules}
\label{app:failures}

\paragraph{The sign factor in straight-through mask gradients.} Selecting the
top-$k$ mask by $|s|$ while passing the straight-through gradient to $s$
directly drops the $\partial|s|/\partial s = \operatorname{sign}(s)$ factor;
every negative-score synapse then moves in the wrong direction, and learning
sits at chance. Applying $|\cdot|$ \emph{before} the straight-through
operator (so autodiff supplies the sign), or equivalently multiplying manual
updates by $\operatorname{sign}(s)$, restores learning entirely. We flag this
because the failure is silent: losses hover near initialization and no error
is raised.

\paragraph{Hebbian runaway and its stabilization.} A naive three-factor rule
diverges here: with ReLU activity the eligibility $a_i x_j$ is non-negative,
so reinforced co-activations grow activations, which grow eligibilities,
until logits saturate (magnitudes ${>}10^2$ within 200 steps), the softmax
collapses to one-hot, exploration dies, and accuracy pins at chance.
Batch-centered (covariance) eligibilities, RMS-normalized per-layer updates,
and a loss-based rather than sampled-action reward stabilize it; the
stabilized rule learns but remains far below the vector-signal rules.

\end{document}
